### 2020-11-18 latex/acrobot.tex
Updated \mathbb{S}^1 to \Sone, made Gamma \simeq \SxR instead of eq.

\section{Dynamics of the Acrobot}
Suppose we are given an acrobot as in Figure \ref{fig:acrobot-model} modelling a
gymnast hanging on a horizontal bar, where the ``torso" has moment of
inertia \(J_u\) and the ``leg" has moment of inertia \(J_a\) (each with respect
to their own center of mass).
Let \(q_u \in \mathbb{S}^1\) be the shoulder angle and \(q_a \in \mathbb{S}^1\) 
be the hip angle, where only \(q_a\) is actuated. 
Collecting them together provides the configuration
\(q = (q_u,q_a) \in \mathbb{S}^1 \times \mathbb{S}^1\). 
The acrobot has inertia matrix \(D\), potential function \(P\) (with respect to
the horizontal bar), and input matrix \(B\)  given as
follows \cite{xingbo_thesis}:
\begin{align}\label{eqn:general-acrobot-inertia}
    D(q) &= \begin{bmatrix}
      m_al_u^2 + 2m_a\cos(q_a)l_u l_{c_a} + m_al_{c_a}^2 + m_ul_{c_u}^2 + J_u + J_a &
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a \\
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a &
      m_al_{c_a}^2 + J_a
    \end{bmatrix} 
    , \\
    \label{eqn:general-acrobot-potential}
    P(q) &= g\left(m_al_{c_a}(1 - \cos(q_u+q_a)) + 
        (m_al_u + m_ul_{c_u})(1-\cos(q_u))\right) 
    , \\
    B(q) &= \begin{bmatrix} 0 \\ 1 \end{bmatrix}
    .
\end{align}

@@

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau,
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
    .
\end{equation}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.
This means \((q,p)\) are simply actuated coordinates inside the phase space
\(\mathcal{Q} \times \mathcal{P}\) where
\(\mathcal{Q} = \mathcal{Q}_u \times \mathcal{Q}_a 
:= \mathbb{S}^1 \times \mathbb{S}^1\), and
\(\mathcal{P} = \mathcal{P}_u \times \mathcal{P}_a
:= \R \times \R\).
This allows us to apply the theory of VNHCs from Chapter
\ref{ch:vnhcs}.

@@

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma = \mathbb{S}^1 \times \R\).
    Let 
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
        ,
    \]
    be the energy of the simple pendulum obtained by setting \(I = 0\).
\begin{enumerate}
    \item There always exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \mathbb{S}^1 \times \R_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\mu_0)d\sigma > 0\) for all 
    \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right[\).
    Then there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(0,\bar{\rho})\right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}
%\footnotetext{In fact, we can be more precise than this. 
%    Letting \(\epsilon > 0\), we need 
%    \(I \leq \min\left\{I_1(\epsilon),I_2\}\) where
%    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
%    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
%    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
%    \right)^{-1/2}\)
%    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
%    See Apendix \ref{apx:I-bounds} for details.
%}
The region \(\mathcal{O}\) is the oscillation domain of the simple pendulum
obtained by setting \(q_a = 0\);
\(\Omega\) includes \(\mathcal{O}\), alongside all rotations of this 
pendulum with momentum less than \(\bar{\mu}\). 
Hence, the first result of Theorem \ref{thm:acrobot-energy-stabilization} states
that the acrobot will gain enough energy to begin performing giant motion;
the second result guarantees can achieve giants with momentum
\(|p_u| < \bar{\mu}\).

@@

There is a suitable change of coordinates into a pseudo-radius 
\(\mu \in \R_{>0}\) and pseudo-angle \(\alpha \in \mathbb{S}^1\) on the
\((q_u,p_u)\)-plane.
These coordinates have the property that \(\dot{\alpha} > 0\) when
\(I\) is small enough.
Perturbation theory shows that (for possibly smaller \(I\)) the radius \(\mu\)
increases on average along \(\alpha\).
Hence, orbits in the \((q_u,p_u)\)-plane spiral away from the origin, 
which means the acrobot is gaining energy.

@@

Using a method developed by \citet{dynamic_vhcs_stabilize_closed_orbits},
we can find a change of coordinates \((q_u,p_u) \to (\alpha, \mu)\) with the
following properties:
\begin{itemize}
    \item The energy of oscillation in \((\alpha,\mu)\) coordinates is 
        uniquely defined by \(\mu\), which remains constant along oscillations
        of the simple pendulum.
    \item \(\alpha\) is a pseudo-angle living in \(\mathbb{S}^1\) which is
        always increasing.
\end{itemize}
Once we have these coordinates in hand, we can use perturbation theory to prove
that \(\mu\) increases on average for the acrobot when \(I\) is small enough.
Likewise, we will find a second set of coordinates \((\beta, \rho)\) where
\(\beta \in \mathbb{S}^1\) is always increasing, \(\rho\) is constant along
rotations of the simple pendulum, and \(\rho\) increases for the acrobot when
\(I\) is small enough.
Putting these together, we will show the acrobot is gaining energy.

@@

\subsection*{Perturbation Analysis for Oscillations}
As we have seen, the acrobot becomes a nominal pendulum that oscillates whenever the
nominal energy ~\eqref{eqn:acrobot-nominal-E} is less than \(E(\pi,0)\).
Hence, the domain of oscillations for the nominal pendulum is given by
\[
    \mathcal{O} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) < E(\pi,0)\right\}
    ,
\]
where \(\Gamma = \mathbb{S}^1\times\R\) is the consraint manifold of the
acrobot.
We wish to find a transformation \(T(q_u,p_u)\) on \(\mathcal{O}\) into
\((\alpha,\mu)\)-coordinates where \(\alpha \in \mathbb{S}^1\) is a pseudo-angle
and \(\mu > 0\) is a pseudo-radius.
Since the energy of oscillation can be determined by the orbit's intersection
with the \(q_u\)-axis (as in Figure \ref{fig:mu-intersection}) 
and \(q_u \in ]-\pi,\pi[\) on \(\mathcal{O}\), we set \(\mu \in ]0,\pi[\).
The transformation we want is therefore a diffeomorphism of the form
\begin{align*}
    T : D\backslash \{(0,0)\} &\rightarrow \mathbb{S}^1 \times \, ]0,\pi[, \\
    (q_u, p_u) &\mapsto (\alpha,\mu)
    .
\end{align*}

@@

\subsection*{Perturbation Analysis for Rotations}
The rotation region for the nominal pendulum is the set
\[
    \mathcal{R} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) > E(\pi,0)\right\}
    .
\]
We wish to find a new set of coordinates \((\beta,\rho)\) for the acrobot
and perform a similar analysis to what we did for oscillations.
In the oscillation region \(\mathcal{O}\) we showed that the nominal pendulum's
energy (and hence, its orbit) is uniquely determined by the intersection point
\(\mu\) with the \(q_u\)-axis; 
when the nominal pendulum is rotating, the energy is instead determined by the
intersection \(\rho\) on the \(p_u\)-axis.
Note that the boundary of \(\mathcal{O}\) intersects the \(p_u\) axis when 
\(p_u^2 = 60m^2 g l^3\); 
hence, for a rotation we must have \(\rho > \sqrt{60 m^2 g l^3}\) if the
orbit has momentum \(p_u > 0\) 
or \(\rho < -\sqrt{60 m^2 g l^3}\) if the orbit has momentum \(p_u < 0\).
The energy level set associated with \(\rho\) is
\[
    \left\{(q_u,p_u) \in \Gamma \mid E(q_u,p_u) = \frac{\rho^2}{10ml^2}\right\}
    ,
\]
which gives the relationship
\begin{equation}\label{eqn:rotation-pu2}
    \frac{p_u^2}{10m l^2} + 30mgl(1 - c_u) = \frac{\rho^2}{10 ml^2}
    .
\end{equation}
On this level set, \(q_u\) takes all values on \(\mathbb{S}^1\), so our angle of
rotation is uniquely parameterized by \(\beta = q_u\).
Since \(\rho\) does not change sign along the rotation, we have the smooth
relationship
\begin{align}\label{eqn:rotation-Tinv}
    q_u &= \beta
    , \\
    p_u &= \sign{\rho}\sqrt{\rho^2 - 30 m^2 g l^3\left(1 - c_\beta \right)}
    .
\end{align}

Inverting this relationship gives our (smooth) transformation
\begin{equation*}
    T : \mathbb{S}^1 \times \R \rightarrow 
    \mathbb{S}^1 \times ]-\infty,-60m^2gl^3[\, \cup \, ]60m^2gl^3,\infty[ 
    ,
\end{equation*}
where \(T(q_u,p_u) = (\beta,\rho)\) is defined by
\begin{align}\label{eqn:rotation-T}
    \beta &= q_u
    , \\
    \rho &= \sign{p_u}\sqrt{p_u^2 + 30 m^2 g l^3 \left(1 - c_u \right)}
    .
\end{align}

>>>>

\section{Dynamics of the Acrobot}
Suppose we are given an acrobot as in Figure \ref{fig:acrobot-model} modelling a
gymnast hanging on a horizontal bar, where the ``torso" has moment of
inertia \(J_u\) and the ``leg" has moment of inertia \(J_a\) (each with respect
to their own center of mass).
Let \(q_u \in \Sone\) be the shoulder angle and \(q_a \in \Sone\) 
be the hip angle, where only \(q_a\) is actuated. 
Collecting them together provides the configuration
\(q = (q_u,q_a) \in \Sone \times \Sone\). 
The acrobot has inertia matrix \(D\), potential function \(P\) (with respect to
the horizontal bar), and input matrix \(B\)  given as
follows \cite{xingbo_thesis}:
\begin{align}\label{eqn:general-acrobot-inertia}
    D(q) &= \begin{bmatrix}
      m_al_u^2 + 2m_a\cos(q_a)l_u l_{c_a} + m_al_{c_a}^2 + m_ul_{c_u}^2 + J_u + J_a &
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a \\
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a &
      m_al_{c_a}^2 + J_a
    \end{bmatrix} 
    , \\
    \label{eqn:general-acrobot-potential}
    P(q) &= g\left(m_al_{c_a}(1 - \cos(q_u+q_a)) + 
        (m_al_u + m_ul_{c_u})(1-\cos(q_u))\right) 
    , \\
    B(q) &= \begin{bmatrix} 0 \\ 1 \end{bmatrix}
    .
\end{align}

@@

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau,
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
    .
\end{equation}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.
This means \((q,p)\) are simply actuated coordinates inside the phase space
\(\mathcal{Q} \times \mathcal{P}\) where
\(\mathcal{Q} = \mathcal{Q}_u \times \mathcal{Q}_a 
:= \Sone \times \Sone\), and
\(\mathcal{P} = \mathcal{P}_u \times \mathcal{P}_a
:= \R \times \R\).
This allows us to apply the theory of VNHCs from Chapter
\ref{ch:vnhcs}.

@@

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma \simeq \SxR\).
    Let 
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
        ,
    \]
    be the energy of the simple pendulum obtained by setting \(I = 0\).
\begin{enumerate}
    \item There always exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \SxR_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\mu_0)d\sigma > 0\) for all 
    \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right[\).
    Then there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
            \mid E(q_u,p_u) < E(0,\bar{\rho})\right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}
%\footnotetext{In fact, we can be more precise than this. 
%    Letting \(\epsilon > 0\), we need 
%    \(I \leq \min\left\{I_1(\epsilon),I_2\}\) where
%    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
%    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
%    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
%    \right)^{-1/2}\)
%    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
%    See Apendix \ref{apx:I-bounds} for details.
%}
The region \(\mathcal{O}\) is the oscillation domain of the simple pendulum
obtained by setting \(q_a = 0\);
\(\Omega\) includes \(\mathcal{O}\), alongside all rotations of this 
pendulum with momentum less than \(\bar{\mu}\). 
Hence, the first result of Theorem \ref{thm:acrobot-energy-stabilization} states
that the acrobot will gain enough energy to begin performing giant motion;
the second result guarantees can achieve giants with momentum
\(|p_u| < \bar{\mu}\).

@@

There is a suitable change of coordinates into a pseudo-radius 
\(\mu \in \R_{>0}\) and pseudo-angle \(\alpha \in \Sone\) on the
\((q_u,p_u)\)-plane.
These coordinates have the property that \(\dot{\alpha} > 0\) when
\(I\) is small enough.
Perturbation theory shows that (for possibly smaller \(I\)) the radius \(\mu\)
increases on average along \(\alpha\).
Hence, orbits in the \((q_u,p_u)\)-plane spiral away from the origin, 
which means the acrobot is gaining energy.

@@

Using a method developed by \citet{dynamic_vhcs_stabilize_closed_orbits},
we can find a change of coordinates \((q_u,p_u) \to (\alpha, \mu)\) with the
following properties:
\begin{itemize}
    \item The energy of oscillation in \((\alpha,\mu)\) coordinates is 
        uniquely defined by \(\mu\), which remains constant along oscillations
        of the simple pendulum.
    \item \(\alpha\) is a pseudo-angle living in \(\Sone\) which is
        always increasing.
\end{itemize}
Once we have these coordinates in hand, we can use perturbation theory to prove
that \(\mu\) increases on average for the acrobot when \(I\) is small enough.
Likewise, we will find a second set of coordinates \((\beta, \rho)\) where
\(\beta \in \Sone\) is always increasing, \(\rho\) is constant along
rotations of the simple pendulum, and \(\rho\) increases for the acrobot when
\(I\) is small enough.
Putting these together, we will show the acrobot is gaining energy.

@@

\subsection*{Perturbation Analysis for Oscillations}
As we have seen, the acrobot becomes a nominal pendulum that oscillates whenever the
nominal energy ~\eqref{eqn:acrobot-nominal-E} is less than \(E(\pi,0)\).
Hence, the domain of oscillations for the nominal pendulum is given by
\[
    \mathcal{O} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) < E(\pi,0)\right\}
    ,
\]
where \(\Gamma \simeq \Sone\times\R\) is the constraint manifold of the
acrobot.
We wish to find a transformation \(T(q_u,p_u)\) on \(\mathcal{O}\) into
\((\alpha,\mu)\)-coordinates where \(\alpha \in \Sone\) is a pseudo-angle
and \(\mu > 0\) is a pseudo-radius.
Since the energy of oscillation can be determined by the orbit's intersection
with the \(q_u\)-axis (as in Figure \ref{fig:mu-intersection}) 
and \(q_u \in ]-\pi,\pi[\) on \(\mathcal{O}\), we set \(\mu \in ]0,\pi[\).
The transformation we want is therefore a diffeomorphism of the form
\begin{align*}
    T : D\backslash \{(0,0)\} &\rightarrow \Sone \times \, ]0,\pi[, \\
    (q_u, p_u) &\mapsto (\alpha,\mu)
    .
\end{align*}

@@

\subsection*{Perturbation Analysis for Rotations}
The rotation region for the nominal pendulum is the set
\[
    \mathcal{R} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) > E(\pi,0)\right\}
    .
\]
We wish to find a new set of coordinates \((\beta,\rho)\) for the acrobot
and perform a similar analysis to what we did for oscillations.
In the oscillation region \(\mathcal{O}\) we showed that the nominal pendulum's
energy (and hence, its orbit) is uniquely determined by the intersection point
\(\mu\) with the \(q_u\)-axis; 
when the nominal pendulum is rotating, the energy is instead determined by the
intersection \(\rho\) on the \(p_u\)-axis.
Note that the boundary of \(\mathcal{O}\) intersects the \(p_u\) axis when 
\(p_u^2 = 60m^2 g l^3\); 
hence, for a rotation we must have \(\rho > \sqrt{60 m^2 g l^3}\) if the
orbit has momentum \(p_u > 0\) 
or \(\rho < -\sqrt{60 m^2 g l^3}\) if the orbit has momentum \(p_u < 0\).
The energy level set associated with \(\rho\) is
\[
    \left\{(q_u,p_u) \in \Gamma \mid E(q_u,p_u) = \frac{\rho^2}{10ml^2}\right\}
    ,
\]
which gives the relationship
\begin{equation}\label{eqn:rotation-pu2}
    \frac{p_u^2}{10m l^2} + 30mgl(1 - c_u) = \frac{\rho^2}{10 ml^2}
    .
\end{equation}
On this level set, \(q_u\) takes all values on \(\Sone\), so our angle of
rotation is uniquely parameterized by \(\beta = q_u\).
Since \(\rho\) does not change sign along the rotation, we have the smooth
relationship
\begin{align}\label{eqn:rotation-Tinv}
    q_u &= \beta
    , \\
    p_u &= \sign{\rho}\sqrt{\rho^2 - 30 m^2 g l^3\left(1 - c_\beta \right)}
    .
\end{align}

Inverting this relationship gives our (smooth) transformation
\begin{equation*}
    T : \SxR \rightarrow 
    \Sone \times ]-\infty,-60m^2gl^3[\, \cup \, ]60m^2gl^3,\infty[ 
    ,
\end{equation*}
where \(T(q_u,p_u) = (\beta,\rho)\) is defined by
\begin{align}\label{eqn:rotation-T}
    \beta &= q_u
    , \\
    \rho &= \sign{p_u}\sqrt{p_u^2 + 30 m^2 g l^3 \left(1 - c_u \right)}
    .
\end{align}
